---
title: "STA 108 Project 1"
author: "Mahir Oza, Dylan M. Ang, In Seon Kim"
date: "1/31/2022"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(
	error = FALSE,
	message = FALSE,
	warning = FALSE,
	echo = FALSE, # hide all R codes!!
	fig.width=5, fig.height=4,#set figure size
	fig.align='center',#center plot
	options(knitr.kable.NA = ''), #do not print NA in knitr table
	tidy = FALSE #add line breaks in R codes
)
```


## Background

The United States Census Bureau is a government organization that collects and produces data about the American people and economy. Its mission is to display quality data about the population in all regions of America. Every 10 years, they conduct a population and housing census that includes all residents living in the United States. They not only count the population, but also gather information about different factors in order to analyze the people and economy.

Throughout our project, we analyze the County Demographic Information (CDI) provided by the United States Census Bureau. The data set displays information about 440 populous counties in the United States with 14 different variables to describe the county's economy. The counties range from all throughout the US, from Orange County in California to Wayne County in North Carolina. Various data variables such as land area, total population, number of active physicians, number of hospital beds, total serious crime, percent high school graduates, percent bachelor's degree, etc were gathered for a single county. Some counties with missing data were deleted from the data set. The data reflects the years 1990 and 1992. 

We will be using R studio, a known professional software used in statistics, in our research to analyze and graph our data. The purpose of this is to demonstrate and analyze a simple linear regression model for our random variables and display inference methods that can be applied to our model. Data is provided from the textbook "Applied Linear Statistical Models" and our project will contain five parts:

- Fitting regression models
- Measuring linear association
- Inference about regression parameters
- Regression diagnostics
- Discussion


## Part 1. Fitting Regression Models

```{r} 
# import data
CDI = read.table("./CDI.txt")

# define active physicians columns
phys = CDI$V8

# define predictor variables total pop, 
#         hospital beds, and personal income 
pop = CDI$V5
bed = CDI$V9
inc = CDI$V16
```

### a) Three Predictors for Active Physicians

All regression functions appear in the form,

\begin{equation}\label{eq: gen_regress}
  Y_i = \hat\beta_0 + \hat\beta_1 X_i + \epsilon_i
\end{equation}

```{r}
# Physicians vs Population
fit_pop = lm(phys ~ pop)
b0_pop = fit_pop$coefficients[1]
b1_pop = fit_pop$coefficients[2]

# Physicians vs Beds
fit_bed = lm(phys ~ bed)
b0_bed = fit_bed$coefficients[1]
b1_bed = fit_bed$coefficients[2]

# Physicians vs Income
fit_inc = lm(phys ~ inc)
b0_inc = fit_inc$coefficients[1]
b1_inc = fit_inc$coefficients[2]
```

\begin{align}
  \text{Active Physicians } & \text{vs. Hospital Beds}         && Y = `r b0_bed` + `r b1_bed`X_i \\
  \text{Active Physicians } & \text{vs. Total Population}      && Y = `r b0_pop` + `r b1_pop`X_i \\
  \text{Active Physicians } & \text{vs. Total Personal Income} && Y = `r b0_inc` + `r b1_inc`X_i
\end{align}

### b) Plots

```{r}
plot(x=pop,y=phys,
     xlab="Total Population", ylab="Number of Active Physicians", 
     main="Active Physicians vs. Total Population")
abline(fit_pop,col="red")

plot(x=bed,y=phys,
     xlab="Number of Hospital Beds", ylab="Number of Active Physicians", 
     main="Active Physicians vs. Number of Hospital Beds")
abline(fit_bed,col="red")

plot(x=inc,y=phys,
     xlab="Total Personal Income (in millions of dollars)", ylab="Number of Active Physicians", 
     main="Active Physicians vs. Total Personal Income")
abline(fit_inc,col="red")
```

### c) MSE Values

```{r}
E_pop = phys - (b0_pop + (b1_pop * pop))
E_bed = phys - (b0_bed + (b1_bed * bed))
E_inc = phys - (b0_inc + (b1_inc * inc))

n_pop = length(pop)
n_bed = length(bed)
n_inc = length(inc)
SSE_p = sum((E_pop)^2)
SSE_b = sum((E_bed)^2)
SSE_i = sum((E_inc)^2)
MSE_p = SSE_p/(n_pop - 2)
MSE_b = SSE_b/(n_bed - 2)
MSE_i = SSE_i/(n_inc - 2)
```

```{r}
sprintf("Population: %f", MSE_p)
sprintf("Beds: %f", MSE_b)
sprintf("Income: %f", MSE_i)
```
The mean of squared errors of the number of 
hospital beds as the predictor for number of active physicians in a county 
is 310,191.883. This is the lower MSE value compared to predictors total personal
income, which had an MSE of 324,539.394, and total population of the county, 
which had the highest MSE of 372,203.505. This shows the mean squared errors in
which the number of hospital beds in a county predicting the number of 
active physicians is lowest and therefore has the lowest variability 
around the fitted regression line.


## Part 2. Measuring Linear Associations

$R^2$ values

```{r}
fit_pop_sum = summary(fit_pop)
fit_bed_sum = summary(fit_bed)
fit_inc_sum = summary(fit_inc)
r2p = fit_pop_sum$r.squared
r2b = fit_bed_sum$r.squared
r2i = fit_inc_sum$r.squared
sprintf("Population: %f", r2p)
sprintf("Beds: %f", r2b)
sprintf("Income: %f", r2i)
```

For predicting the number of active physicians in a county, 
the $R^2$ value for total population as the predictor is 0.8841, for 
number of hospital beds is 0.9034, and for total personal income is 
0.8989. Based on these values, the number of hospital beds as the predictor 
has the largest coefficient of determination meaning that 90.34% of the 
variation in the number of active physicians is due to introducing 
number of hospital beds into the regression model. This shows that the 
number of hospital beds accounts for the largest reduction in variability 
in the number of active physicians.

## Part 3. Inference About Regression Parameters

Confidence Intervals for per capita income (y) vs percent of individuals
with atleast bachelor's degree (x) for each region
```{r}

# split region
# V12: % Bachelors and V15: per capita income
region = CDI$V17 == "1"
NE = CDI[region, c("V12", "V15")]

region = CDI$V17 == "2"
NC = CDI[region, c("V12", "V15")]

region = CDI$V17 == "3"
S = CDI[region, c("V12", "V15")]

region = CDI$V17 == "4"
W = CDI[region, c("V12", "V15")]

# linear models
fit_NE = lm(NE$V15 ~ NE$V12)
fit_NC = lm(NC$V15 ~ NC$V12)
fit_S = lm(S$V15 ~ S$V12)
fit_W = lm(W$V15 ~ W$V12)

# betas
b0_NE = fit_NE$coefficients[1]
b0_NC = fit_NC$coefficients[1]
b0_S = fit_S$coefficients[1]
b0_W = fit_W$coefficients[1]

b1_NE = fit_NE$coefficients[2]
b1_NC = fit_NC$coefficients[2]
b1_S = fit_S$coefficients[2]
b1_W = fit_W$coefficients[2]

# SSEs
SSE_NE = sum( ( NE$V15 - (b0_NE + b1_NE * NE$V12) )^2 )
SSE_NC = sum( ( NC$V15 - (b0_NC + b1_NC * NC$V12) )^2 )
SSE_S = sum( ( S$V15 - (b0_S + b1_S * S$V12) )^2 )
SSE_W = sum( ( W$V15 - (b0_W + b1_W * W$V12) )^2 )

# MSEs
MSE_NE = SSE_NE / (nrow(NE) - 2)
MSE_NC = SSE_NC / (nrow(NC) - 2)
MSE_S = SSE_S / (nrow(S) - 2)
MSE_W = SSE_W / (nrow(W) - 2)

# t vals 
alpha = 0.1
t_NE = qt(1 - alpha/2, nrow(NE) - 2)
t_NC = qt(1 - alpha/2, nrow(NC) - 2)
t_S = qt(1 - alpha/2, nrow(S) - 2)
t_W = qt(1 - alpha/2, nrow(W) - 2)

# SEs
SE_NE = sqrt( MSE_NE / sum( (NE$V12 - mean(NE$V12) )^2) )
SE_NC = sqrt( MSE_NC / sum( (NC$V12 - mean(NC$V12) )^2) )
SE_S = sqrt( MSE_S / sum( (S$V12 - mean(S$V12) )^2) )
SE_W = sqrt( MSE_W / sum( (W$V12 - mean(W$V12) )^2) )

#  CIs
NE_up = b1_NE + t_NE * SE_NE
NE_lo = b1_NE - t_NE * SE_NE

NC_up = b1_NC + t_NC * SE_NC
NC_lo = b1_NC - t_NC * SE_NC

S_up = b1_S + t_S * SE_S
S_lo = b1_S - t_S * SE_S

W_up = b1_W + t_W * SE_W
W_lo = b1_W - t_W * SE_W

# print CIs
sprintf("NE(1): [%.3f, %.3f]", NE_lo, NE_up)
sprintf("NC(2): [%.3f, %.3f]", NC_lo, NC_up)
sprintf("S (3): [%.3f, %.3f]", S_lo, S_up)
sprintf("W (4): [%.3f, %.3f]", W_lo, W_up)
```
The regression lines for all 4 regions seem to have pretty large, positive
slopes based on their 90% confidence intervals where the lower
limit of each interval seems to 
be greater that 190. However the magnitudes of these large, positive slopes and
even the size of the confidence intervals differ slightly based on
region. The NC region
has an interval of [193.486, 283.853], the S region has an interval of 
[285.708, 375.516], the W region has an interval of [364.758,515,873], and 
the NE region has an interval of [460.518,583.800]. These intervals represent
that we are 90% confident the true slope of per capita income vs percent
of individuals with atleast a bachelor's degree lies in the interval for each
respective region.

```{r}
# SSRs
SSR_NE = sum( ( (b0_NE + b1_NE * NE$V12) - mean(NE$V15) )^2 )
SSR_NC = sum( ( (b0_NC + b1_NC * NC$V12) - mean(NC$V15) )^2 )
SSR_S = sum( ( (b0_S + b1_S * S$V12) - mean(S$V15) )^2 )
SSR_W = sum( ( (b0_W + b1_W * W$V12) - mean(W$V15) )^2 )

# MSRs
MSR_NE = SSR_NE
MSR_NC = SSR_NC
MSR_S = SSR_S
MSR_W = SSR_W

# F-statistics
F_NE = MSR_NE/MSE_NE
F_NC = MSR_NC/MSE_NC
F_S = MSR_S/MSE_S
F_W = MSR_W/MSE_W
```
NE F-test: per capita income (y) vs percent of individuals
with atleast bachelor's degree (x)

$$H_0: \beta_1 = 0$$
$$H_1: \beta_1 \neq 0$$
```{r}
p_NE = 1-pf(F_NE,1,nrow(NE)-2)
```
\begin{align}
  F = `r F_NE` \\
  \text{p-value} = `r p_NE` 
\end{align}
For significance level $\alpha = 0.01$ the F-statistic for the NE region, 
197.753, shows an incredibly small p-value where
p-value $= 0 < 0.01 = \alpha$, 
meaning that $H_0$ is rejected. This conclusion as well as the fact that the 
F-statistic is very large, shows that regression for per capita income vs
percent of individuals with a bachelor's degree in the NE region is useful.

NC F-test: per capita income (y) vs percent of individuals
with atleast bachelor's degree (x)

$$H_0: \beta_1 = 0$$
$$H_1: \beta_1 \neq 0$$
```{r}
p_NC = 1-pf(F_NC,1,nrow(NC)-2)
```
\begin{align}
  F = `r F_NC` \\
  \text{p-value} = `r p_NC` 
\end{align}
For significance level $\alpha = 0.01$ the F-statistic for the NC region, 
76.826, shows a small p-value such that
p-value $=$ `r p_NC` $< 0.01 = \alpha$, 
meaning that $H_0$ is rejected. This conclusion as well as the fact that the 
F-statistic is fairly large, shows that regression for per capita income vs
percent of individuals with a bachelor's degree in the NC region is useful.

S F-test: per capita income (y) vs percent of individuals
with atleast bachelor's degree (x)

$$H_0: \beta_1 = 0$$
$$H_1: \beta_1 \neq 0$$
```{r}
p_S = 1-pf(F_S,1,nrow(S)-2)
```
\begin{align}
  F = `r F_S` \\
  \text{p-value} = `r p_S` 
\end{align}
For significance level $\alpha = 0.01$ the F-statistic for the S region, 
148.491, shows a very small p-value where
p-value $= 0 < 0.01 = \alpha$, 
meaning that $H_0$ is rejected. This conclusion as well as the fact that the 
F-statistic is very large, shows that regression for per capita income vs
percent of individuals with a bachelor's degree in the S region is useful.

W F-test: per capita income (y) vs percent of individuals
with atleast bachelor's degree (x)

$$H_0: \beta_1 = 0$$
$$H_1: \beta_1 \neq 0$$
```{r}
p_W = 1-pf(F_W,1,nrow(W)-2)
```
\begin{align}
  F = `r F_W` \\
  \text{p-value} = `r p_W` 
\end{align}
For significance level $\alpha = 0.01$ the F-statistic for the W region, 
94.195, shows a small p-value such that
p-value $=$ `r p_W` $< 0.01 = \alpha$, 
meaning that $H_0$ is rejected. This conclusion as well as the fact that the 
F-statistic is fairly large, shows that regression for per capita income vs
percent of individuals with a bachelor's degree in the W region is useful.


## Part 4. Regression Diagnostics
### Plot 1: Number of Active Physicians (Y) vs Total Population (Xp)
```{r}
#Residual Plot 
plot(CDI$V5, fit_pop$residuals, 
     xlab="Total Population", ylab="Residuals",
     main="Total Population and Residuals")
abline(h=0, col = 'red')

#Normal Probability Plot
qqplot = qqnorm(fit_pop$residuals)
qqline(fit_pop$residuals, col='red')
```

### Plot 2: Number of Active Physicians (Y) vs Number of Hospital Beds (Xh)
```{r}
#Residual Plot
plot(CDI$V9, fit_bed$residuals, 
     xlab="Hospital Beds", ylab="Residuals",
     main="Total Hospital Beds and Residuals")
abline(h=0, col='red')

#Normal Probability Plot
qqplot = qqnorm(fit_bed$residuals)
qqline(fit_bed$residuals, col='red')
```

### Plot 3: Number of Active Physicians (Y) vs Total Personal Income (Xi)
```{r}
plot(CDI$V16, fit_inc$residuals,
     xlab="Total Personal Income (in millions of dollars)", ylab = "Residuals",
     main="Total Personal Income and Residuals")
abline(h=0, col='red')

#Normal Probability Plot
qqplot = qqnorm(fit_inc$residuals)
qqline(fit_inc$residuals, col='red')
```





The residual plot for all 3 predictor variables as a 
function of the number of active physicians shows no distinguishable 
mathematical trend nor any pattern. Also the sum of all the residuals for the 3
predictor variables are extremely small values, close to 0. These show
that a linear regression model is appropriate in showing the relationship 
of the data. In addition to these arguments for a linear regression 
relationship
being a good fit for the data, we can see that the data itself appears to 
demonstrate a linear relationship. Adding the regression function to the data
further shows that the data of all 3 predictor variables follows the behavior
of this regression function and further supports a linear regression 
relationship for all 3 predictor variables for the number of active physicians.
While all 3 QQ plots demonstrate a similar form, it seems that the QQ
plot for the number of active physicians vs number of hospital beds, shows
the best linear trend in the plot compared to the other 2 predictor variables. 
This helps shows that this predictor variable is slighlty more appropriate
in determining a regression relationship for number of active physicians.


## Part 5. Discussion


## Appendix

```{r, ref.label=knitr::all_labels(), echo=TRUE, eval=FALSE}
```










